---
title: "Rideshare & Bus Ride Exploratory Analysis"
author: "Team 16"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Rideshare

### Data Exploratory

This part aims to look for specific pattern from the [Austin rideshare dataset](https://data.world/andytryba/rideaustin) to identify possible features to be used for regression model analysis.

**Instruction**: Download the "RideAustin_Weather" csv file from the link location and save that under "Data" folder. The code file is to be downloaded into "Code" folder. 

Set Working Directory to the current directory under "Code" folder path

#### Call needed libarady and read data into a dataframe

```{r datasetup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(plotly)
library(gridExtra)
library(tidyverse)
library(lubridate)
```

```{r read_data}
rideshare <- read.csv('../Data/RideAustin_Weather.csv')

summary(rideshare)
```

### Data Wrangling & Transformation

From the summary output above, column **distance_traveled** represent the ride distance in meters. The biggest value amount to 400km+, which could be data mistakes. Looking at the quantile, 75% of the data fall below 11km. An assumption is made that passengers would not consider public transportation if the travel distance is >15km.

Hence, data is limited to rides <15km distance traveled. For the easy of view, converting distance_traveled into a new column dist_km by dividing value by 1000. Visualize the distribution of distance traveled by km.


```{r filter, echo=FALSE}
rideshare_filter <- rideshare %>% filter(distance_travelled < 15000)
rideshare_filter <- rideshare %>% filter(distance_travelled < 20000)
rideshare_filter$dist_km <- rideshare_filter$distance_travelled/1000

dist_travel <- ggplot(rideshare_filter, aes(dist_km)) + geom_histogram(binwidth = 1) +
  labs(title = "Distance Distribution by ride share trips",
    x = "Distance Traveled (km)", y = "# of Trips")

dist_travel_cum <- ggplot(rideshare_filter, aes(dist_km)) + stat_ecdf(geom = 'step') +
  labs(title = "Cumlative % for ride share trips",
       x = "Distance Traveled (km)", y = "% to total")

grid.arrange(dist_travel, dist_travel_cum, ncol=2)

```

75% of the trips are within 10km. Most bus services probably has route that covers such distance. The question then become why wouldn't a passenger choose to take public transportation.

Few factors are being considered as follow:

* Time of the day
* Travel duration (assume passengers tolerance of traveled time is < 2 hours)
* Day of the week
* Weather conditions (if impact Day distribution)

```{r feature, echo=FALSE, warning = FALSE}
# Obtain day of week
rideshare_filter$dow <- wday(ymd(rideshare_filter$Date), label = TRUE)
# Obtain month
rideshare_filter$mo <- month(ymd(rideshare_filter$Date), label = TRUE)

# Convert start and end time text into datetime object

rideshare_filter$start_ts <- as.POSIXct(rideshare_filter$started_on,
  tz = Sys.timezone())
rideshare_filter$end_ts <- as.POSIXct(rideshare_filter$completed_on,
  tz = Sys.timezone())

# Obtain hour of the day
rideshare_filter$hour <- rideshare_filter$start_ts %>%
  hour()

# Calculate time traveled
rideshare_filter$ts_travel <- difftime(rideshare_filter$end_ts,
  rideshare_filter$start_ts, units = "mins")

# Limit travel time to be less than 120 minutes
rideshare_ts <- rideshare_filter %>%
  filter(ts_travel > 0 & ts_travel < 120)


dist_ts <- ggplot(rideshare_ts, aes(ts_travel)) + stat_ecdf(geom = "step") +
  labs(title = "Time Traveled by ride share trips",
       x = "Time (mins)", y = "% to total")

day_count <- ggplot(rideshare_filter, aes(x = dow)) + geom_bar() +
  labs(title = "Ride share trips distribution by Day",
       x = "Day of Week", y = "# Trip")


weather_fog <- day_count + facet_grid(vars(Fog)) +
  labs(title = "Day distribution conditioned on Fog",
       x = "Day", y = "# Trip")

weather_thunder <- day_count + facet_grid(vars(Thunder)) +
  labs(title = "Day distribution conditioned on Thunder",
       x = "Day", y = "# Trip")

hour_count <- ggplot(rideshare_filter, aes(x = hour)) + geom_bar() +
  labs(title = "Time of Day",
       x = "Hour", y = "# of Trip")


grid.arrange(hour_count, dist_ts, day_count, ncol=2, nrow=2)

grid.arrange(weather_fog, weather_thunder, ncol=1, nrow=2)
```

### Observations

* Ride share tend to be more populated around early morning time and late evening
  * Hypothesis to Test: public transportation service hours impact passenger choice
* Almost 90% of rideshare complete in < 25 minutes
  * Hypothesis to Test: travel time in public transportation impact passenger choice
* Ride share happen more frequently over weekend
  * Hypothesis to Test: public transportation frequency during weekend impact passenger choice
  * Hypothesis to Test: ride share over weekend is more cost effective in respect to time traveled than public transportation
* Do not observe significant distribution with/without weather conditions
  * Hypothesis to Test: Passenger preference to use rideshare is not impacted by weather conditions
  
## Bus Ride

Similiar explorations is done on the CapMetro ridership volume [dataset](https://data.texas.gov/Transportation/CapMetro-Shapefiles-AUGUST-2017/5d4c-snum)
.

**Instruction**: After downloading the shape file (ths analysis uses August 2017 data) save that under "Data" folder. 

#### Read Data into dataframe
```{r read_bus}
capmetro_dow_hour <- read.csv("../Data/CapMetro ridership all by latlong dow hour.csv")

```

#### Perform similiar Visualization as shareride

* Group bus ride volume group by day of week and hour of day
* Sum riders volume based on get_on bus metric to obtain total passengers
* Estimate bus freqency by assuming one row is one bus count

```{r group, echo=FALSE, warning = FALSE}
bus_dow <- capmetro_dow_hour %>%
  group_by(start_time_dayofwk) %>%
  summarize(ride = sum(rides_on), n = n()) %>%
  mutate(dow = wday(start_time_dayofwk, label = TRUE))

dow_ride <- ggplot(bus_dow, aes(dow, ride)) + geom_bar(stat = "identity") +
  labs(title = "Ride Volume by Day", x = "Day of Week", y = "# Passengers")

dow_count <- ggplot(bus_dow, aes(dow, n)) + geom_bar(stat = "identity") +
  labs(title = "Est. Bus Count by Day", x = "Day of Week", y = "Est. Bus Count")

bus_hr <- capmetro_dow_hour %>%
  group_by(start_time_hour) %>%
  summarize(ride = sum(rides_on), n = n()) %>% mutate

hr_ride <- ggplot(bus_hr, aes(start_time_hour, ride)) + geom_bar(stat = "identity") +
  labs(title = "Ride Volume by Hour", x = "Hour", y = "# Passengers")

hr_count <- ggplot(bus_hr, aes(start_time_hour, n)) + geom_bar(stat = "identity") +
  labs(title = "Est. Bus Count by hour", x = "Hour", y = "Est. Bus Count")

grid.arrange(dow_ride, dow_count, ncol = 2)
grid.arrange(hr_ride, hr_count, ncol = 2)
```

### Observations

* Bus rides occur more often during weekdays vs weekend, even though the estimated number of bus count are not significant different among day of week
* Bus ride occur mostly from 7am to 4pm. Ridership drop drastically till mid-night even though bus frequency stay relative the same throughout the day
* Minimal bus rides from mid-night to 5am, probably driven by reduce in bus frequency

### Anticipated Hypothesis

* Control on day of week, hour of day, and weather condition, higher bus ride volume lead to lower ride share volume
* For the same from and to location, the longer the difference in time among ride share and public transposition, the higher the ride share volume